---
title: "sampler package"
author: "Bruno Vilela"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
The sampler package generates aggregated or overdispersed samples from any given phylogenetic tree, spatial points defined by two coordinates, or generic distance matrix. Resulting samples can remove data bias in existing datasets or help design sampling schemes for new experiments.
  
The package includes 4 functions: 

* run_sampler: Generates aggregated (underdispersed) or overdispersed samples of values from any given distance matrix. 

* run_sampler_geo: Generates aggregated (underdispersed) or overdispersed samples of points from any given set of geographic coordinates. 

* run_sampler_phy: Generates aggregated (underdispersed) or overdispersed samples of tips from any given phylogenetic tree. 

* Nee_May_1997: Generates a subsample of tips from a phylogenetic tree that maximize the remaining amount of evolutionary history for each node (see Nee and May 1997).

  
  
## run_sampler
  
Phylogeny example:
```{r, message = FALSE}
require(sampler)
require(ape)
```

```{r, , fig.width=6, fig.height=6}
## Generate a random tree
tree <- rcoal(10)
## Calculate the distance
dist <- cophenetic(tree)
## Highly overdispersed 50% resample design (alpha = 50)
selection <- run_sampler(x = dist, n = 5, alpha = 50, starting = "t10")
## Highly aggregated 50% resample design (alpha = -50)
selection2 <- run_sampler(x = dist, n = 5, alpha = -50, starting = "t10")
## Random 50% resample design (alpha = 0)
selection3 <- run_sampler(x = dist, n = 5, alpha = 0, starting = "t10")
## Plot to compare

plot(tree,tip.color=ifelse(tree$tip.label %in% selection, "red","black"),
     main = "Overdispersed 50% sampling (red were selected)", cex = 1)
axis(1)
plot(tree,tip.color=ifelse(tree$tip.label %in% selection2, "blue","black"),
     main = "Aggregated 50% sampling (blue were selected)", cex = 1)
axis(1)
plot(tree,tip.color=ifelse(tree$tip.label %in% selection3, "green","black"),
     main = "Random 50% sampling (green were selected)", cex = 1)
axis(1)
```

Geography example:
```{r, message = FALSE}
require(sp)
require(maptools)
require(fields)
```

```{r, fig.width=6, fig.height=6}
data(wrld_simpl)  # World map
Brazil <- wrld_simpl[wrld_simpl$NAME == "Brazil", ]  # Brazil (polygon)
coords <- slot(spsample(Brazil, 100, "regular"), "coords")
rownames(coords) <- paste0("t", 1:nrow(coords))
## Calculate the geographic distance
dist.geo <- rdist.earth(coords)
## Subsample 50%
### Overdispersed
selection.geo <- run_sampler(x = dist.geo, n = 25, alpha = 100, starting = "t10")
### Aggregated
selection.geo2 <- run_sampler(x = dist.geo, n = 25, alpha = -100, starting = "t10")
### Random
selection.geo3 <- run_sampler(x = dist.geo, n = 25, alpha = 0, starting = "t10")

## Plot
plot(Brazil, main = "Overdispersed 50% sampling (red were selected)")
points(coords, cex = 2, pch = 19,
       col = ifelse(rownames(coords) %in% selection.geo, "red","gray"))
plot(Brazil, main = "Aggregated 50% sampling (blue were selected)")
points(coords, cex = 2, pch = 19,
       col = ifelse(rownames(coords) %in% selection.geo2, "blue","gray"))
plot(Brazil, main = "Random 50% sampling (green were selected)")
points(coords, cex = 2, pch = 19,
       col = ifelse(rownames(coords) %in% selection.geo3, "green","gray"))
```

Trait example:
```{r, fig.width=6, fig.height=6}
## Fake body size
set.seed <- 1
body_size <- runif(1000)
# Biased sample towards large species
set.seed <- 1
body_size_bias <- sample(body_size, 500, prob = body_size)
hist(body_size, main = "Species body size distribution\n(n = 1000)",  xlab = "Body size")
hist(body_size_bias, main = "Biased samplig towards larger species\n(n = 500)",
xlab = "Body size")
# Use sampler to reduce the bias
dist_bs <- as.matrix(dist(body_size_bias))
rownames(dist_bs) <- colnames(dist_bs) <- 1:length(body_size_bias)
selection.bs <- run_sampler(x = dist_bs, n = 100, alpha = 100)
hist(body_size_bias[as.numeric(selection.bs)],
 main = "Overdispersed sampling of biased information \n(n = 100)",
 xlab = "Body size")
```

## run_sampler_geo
```{r, message= FALSE}
require(sp)
require(maptools)
```

```{r, fig.width=6, fig.height=6}
data(wrld_simpl)  # World map
Brazil <- wrld_simpl[wrld_simpl$NAME == "Brazil", ]  # Brazil (polygon)
coords <- slot(spsample(Brazil, 100, "regular"), "coords")
rownames(coords) <- paste0("t", 1:nrow(coords))
## Subsample 50%
### Overdispersed
selection.geo <- run_sampler_geo(x = coords, n = 10, alpha = 100, starting = "t10")
### Aggregated
selection.geo2 <- run_sampler_geo(x = coords, n = 10, alpha = -100, starting = "t10")
### Random
selection.geo3 <- run_sampler_geo(x = coords, n = 10, alpha = 0, starting = "t10")

## Plot
plot(Brazil, main = "Overdispersed 50% sampling (red were selected)")
points(selection.geo, cex = 2, pch = 19, col = "red")
plot(Brazil, main = "Aggregated 50% sampling (blue were selected)")
points(selection.geo2, cex = 2, pch = 19, col = "blue")
plot(Brazil, main = "Random 50% sampling (green were selected)")
points(selection.geo3, cex = 2, pch = 19, col = "green")
```

## run_sampler_phy

```{r, message = FALSE}
require(ape)
```

```{r, fig.width=6, fig.height=6}
# Generate a random tree
set.seed(100)
tree <- rcoal(10)
set.seed(2)
# Highly overdispersed 50% resample design (alpha = 100)
overdispersed <- run_sampler_phy(tree, 5, alpha = 100, starting = "t10")
# Highly aggregated 50% resample design (alpha = -100)
aggregated <- run_sampler_phy(tree, 5, alpha = -100, starting = "t10")
# Random 50% resample design (alpha = 0)
random <- run_sampler_phy(tree, 5, alpha = 0, starting = "t10")
# Plot to compare
par(mfrow = c(2, 2))
plot(tree, main = "Full tree", cex = 1)
axis(1)
plot(overdispersed, main = "Overdispersed 50% sampling", cex = 1)
axis(1)
plot(aggregated, main = "Aggregated 50% sampling", cex = 1)
axis(1)
plot(random, main = "Random 50% sampling", cex = 1)
axis(1)
```

## Nee_May_1997

```{r, message = FALSE}
require(ape)
```

```{r, fig.width=6, fig.height=6}
set.seed(50)
tree <- rcoal(50)
k = 10
plot(tree, main = paste("Original tree \n n =", Ntip(tree)))
plot(Nee_May_1997(tree, k), main = paste("Tree with phylogenetic history maximized
n =",  k,  "(Nee & May 1997)"))

# Compare the sampler algorithm with Nee and May (1997) method:
PD_ob <- sum(Nee_May_1997(tree, k)$edge.length)
n_a = 20
alphas <- seq(-5, 5, length.out = n_a)
rep = 100
PD <- matrix(ncol = n_a, nrow = rep)
for (j in 1:rep) {
 for (i in 1:n_a) {
 PD[j, i] <- sum(run_sampler_phy(tree, k, alpha = alphas[i], starting = "t10")$edge.length)
 }
}
PD_max <- apply(PD, 2, max)
PD_min <- apply(PD, 2, min)
PD_mean <- apply(PD, 2, mean)

plot(alphas, PD_mean, ylab = "Phylogenetic Diveristy",
 ylim = c(min(PD), max(c(PD, PD_ob))), type = "l", lwd = 2, col = "black")
 lines(alphas, PD_max, lty = 2, col = "gray20")
 lines(alphas, PD_min, lty = 2, col = "gray20")
 points(max(alphas) + max(alphas) *.05, PD_ob, col = "red", pch = 18, cex = 2)
 legend(x = max(alphas) - max(alphas) * .90, y = min(PD) + min(PD) * 2,
       legend = c("Sampler (mean)", "Sampler (min and max)", "Nee & May"),
       pch = c(NA, NA, 18), lty = c(1, 2, NA),
       col = c("black", "gray20","red"))

```

